---
title: "ST662 Group Project"
author: "Susan Edgeworth, Pengyu Yang, Jack Francis Hickey, Aaron John Doyle, James Doherty Ferris"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction 






## Cleaning and Packages
```{r echo=FALSE, warning=FALSE,message=FALSE}
library(nycflights13)
library(dplyr)
library(ggplot2)
library(plotly)
library(RColorBrewer)
library(tidyverse)
library(sf)
library(hrbrthemes)
library(viridis)
library(albersusa)
library("maps")
library("tools")
library(corrplot)
library(kableExtra)

weather <- nycflights13::weather


```

## Methods






## Results and Analysis

## Airline delays and Analysis
The first plot shows the average total delay (Arrival delay + Departure Delay) of each airline. Alaska airlines appears to be the most timely airline compared with others,more analysis needs to be done here to assess the time delays as there are more flights recorded for certain airlines and apparent outliers which may scew the average. 


The plot on the right shows where the total delay is more than an hour. Alaska Airlines still comes out in the top three with 56 flights that had over an hour delay, compared to ExpressJet Airlines with 11503 flights.
```{r echo=FALSE, message=FALSE, warning=FALSE, figures-side, fig.show="hold", out.width="50%"}

#This is to get the total delay which will be used to see what airline has the most and least delays:
flights<- mutate(flights, totaldelay = arr_delay + dep_delay)

# Adding names of the carrier and changing the column name for clarity:
flights<- flights %>% left_join(airlines, by = c('carrier' = 'carrier'))
colnames(flights)[21]<- "CarrierName"

delays<-flights %>% group_by(CarrierName) %>% 
  summarise(Average_Delay = mean(totaldelay, na.rm =
TRUE)) %>% 
  ggplot(aes(reorder(CarrierName,Average_Delay), Average_Delay, fill= CarrierName)) + geom_bar(stat ='identity', color = rainbow(16)) + scale_x_discrete(guide = guide_axis(angle = 90)) 

delays<-delays + theme_classic()+ ggtitle('Average Delay by Airline')+
  xlab("Airline") + ylab("Average Delay")
delays



hourdelay<-flights %>% filter(totaldelay > 60) %>% count(CarrierName, sort =TRUE) %>% 
  mutate(CarrierName = factor(CarrierName, levels = CarrierName, ordered =TRUE)) %>%
ggplot(aes(CarrierName, n, fill = CarrierName)) + geom_bar(stat ='identity', color = rainbow(16)) + scale_x_discrete(guide = guide_axis(angle = 90)) 

hourdelay<-hourdelay + theme_classic()+ ggtitle('Delay by more than 60 Minutes')+
  xlab("Airline") + ylab("Count")
hourdelay

```

This plot shows a count of all 
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=2,fig.align='center', out.width="50%"}
flightsx <- left_join(flights,weather) %>%
  dplyr::select(day, month,dep_delay, arr_delay, carrier,origin, dest, air_time, distance ,time_hour, temp:visib) %>%
  subset(select= -(wind_gust)) #rm wind_gust cause missing too many values
# add time_status and duration columns
flightsx$time_status <- NA
flightsx1 <- flightsx %>% filter(dep_delay > 15) %>% mutate(time_status="delayed")
flightsx2 <- flightsx %>% filter(dep_delay <= 15) %>% mutate(time_status="on_time")
flightsx3<-rbind(flightsx1,flightsx2)
flightsx3 <- flightsx3 %>% arrange(day,month,time_hour)
flightsx3$duration <- NA
flightsx12 <- flightsx3 %>% filter(air_time < 180) %>% mutate(duration="short haul")
flightsx22 <- flightsx3 %>% filter(air_time >= 180 & air_time < 360) %>% mutate(duration="medium haul")
flightsx32 <- flightsx3 %>% filter(air_time >= 360) %>% mutate(duration="long haul")
flightsx4<-rbind(flightsx12,flightsx22,flightsx32)

ggplot(data = flightsx4, aes(x = carrier , fill = time_status)) + labs(title = 'Count of delayed/on-time flights by carrier') + geom_bar(position = position_dodge(preserve = "single")) + theme_classic()+ ggtitle('Delay and On time counts for each airline') + xlab("Airline") + ylab("Count")





```



## Visualising Airport Delays

 This map looks at the airports that have the most delays on arrival, the larger dot represents a bigger delay (in minutes).
https://rpubs.com/suedge12/757742
```{r echo=FALSE,warning=FALSE,message=FALSE,fig.width= 2, fig.height = 2, fig.align='left', fig.show='hold'}
library(albersusa)
m<-usa_sf()
library("maps")
states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
states <- cbind(states, st_coordinates(st_centroid(states)))
library("tools")
states$ID <- toTitleCase(states$ID)


flights1 <- flights %>% group_by(dest) %>% 
  summarise(avg_delay = mean(totaldelay, na.rm = T)) %>% 
  left_join(airports, c("dest"="faa")) %>%
  arrange(desc(avg_delay))



map1<-ggplot(data = m) +
    geom_sf(fill = "chartreuse1",alpha = 0.45) +
    geom_sf(data = states, alpha = 0.25, color = "darkgreen", size = 0.3) + 
    geom_text(data = states, aes(X, Y, label = ID),check_overlap = TRUE,fontface = "bold", size = 2) +
  geom_point(data = flights1, aes(x = lon, y = lat, size = avg_delay, fill = name), color = "purple", alpha = 0.7)+ guides(fill = FALSE) + scale_size(range = c(0, 5), name="Average Delay in Minutes") +
  coord_sf(
    xlim = c(-130, -70),
    ylim = c(20, 50)
  )


map1<-map1 + theme(
  panel.border = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_rect(colour= "aliceblue", fill = "aliceblue",
                                size = 0.5, linetype = "solid"),
  axis.line = element_line(colour = "black"))+ 
  ggtitle("Average Delays by Airport") +
  xlab("Lat") + ylab("Long")
map12<-ggplotly(hide_legend(map1))


```

## Ranking of Airports
```{r echo=FALSE,warning=FALSE,message=FALSE}
top10<- flights1 %>%
  group_by(name) %>%
  arrange(avg_delay) %>%
  top_n(n = 10)


top <- (head(top10[1:3], 6, table.attr = "style='width:30%;'"))

kable(top)%>%
  kable_styling(latex_options = c("HOLD_position", "font_size = 2"))

```

## Potential Delay causes
```{r echo=FALSE, message=FALSE, warning=FALSE}
flightsx <- left_join(flights,weather) %>%
  dplyr::select(day, month,dep_delay, arr_delay, carrier,origin, dest, air_time, distance ,time_hour, temp:visib) %>%
  subset(select= -(wind_gust)) #rm wind_gust cause missing too many values
# add time_status and duration columns
flightsx$time_status <- NA
flightsx1 <- flightsx %>% filter(dep_delay > 15) %>% mutate(time_status="delayed")
flightsx2 <- flightsx %>% filter(dep_delay <= 15) %>% mutate(time_status="on_time")
flightsx3<-rbind(flightsx1,flightsx2)
flightsx3 <- flightsx3 %>% arrange(day,month,time_hour)
flightsx3$duration <- NA
flightsx12 <- flightsx3 %>% filter(air_time < 180) %>% mutate(duration="short haul")
flightsx22 <- flightsx3 %>% filter(air_time >= 180 & air_time < 360) %>% mutate(duration="medium haul")
flightsx32 <- flightsx3 %>% filter(air_time >= 360) %>% mutate(duration="long haul")
flightsx4<-rbind(flightsx12,flightsx22,flightsx32)

```

Look for any possible contributing factors towards these delays:
So by grouping by the three Origin airports for New York (Newark, JFK and La Guardia) and calculating the average values for the weather data it is possible to test correlations between these values and determine contributing weather factors towards departure delays from these airports. Also calculated is the correlation between the number of flights leaving each airport with the departure delays.
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=6, fig.height=3,  out.width="50%"}
weather <- nycflights13::weather
df<- weather %>% left_join(flights) %>%
  filter_at(vars(dep_delay,temp,wind_speed,precip,pressure,visib), all_vars(!is.na(.))) %>%
  group_by(origin) %>%
  summarise(avg_delay = mean(dep_delay), avg_temp = mean(temp), avg_wind = mean(wind_speed),
            avg_precip = mean(precip), avg_pres = mean(pressure), avg_visib = mean(visib), n=n())

library(corrplot)
cors <- cor(df[,-1])
corrplot(cors, type="upper")
```
In terms of average delays per flight Newark is almost 3 minutes more delayed than JFK who in turn is about 2 minutes per flight worse off than La Guardia. A correlation plot is used in order to uncover causes of this.The plot above shows a very strong correlation between number of flights and average delays (~0.979). The weather values dont possess such strong correlations, precipitation being the next strongest (~0.61). No New York airport stands out in terms of more adverse weather conditions, to be expected given their proximity to each each (within 20 miles).


Check which NY airport has best delays record
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=6, fig.height=2.5, out.width="50%"}
a1 <-flightsx4 %>% group_by(origin,time_status) %>%
  summarise(n=n())
ggplot(data=a1,aes(x=origin,y=n,fill=origin)) +
  geom_col() +
  facet_wrap(~time_status) + theme_classic()+
  ggtitle("Number of flights delayed/on-time")+ xlab("Origin") + ylab("Count")
```

In order to determine if excessive flight numbers have an effect on delays below average delays by month is calculated and compared with flights numbers by month, these are split by the three NY airports
```{r echo=FALSE, message=FALSE, warning=FALSE,fig.width=15, fig.height=4}

data <- flights %>% 
  dplyr::select(origin, month, day ,arr_delay, dep_delay) %>%   
  group_by(origin, month) %>% 
  summarise(avg_delay =  mean(dep_delay, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(-avg_delay)
plot1<-ggplot(data, aes(x=month, y=avg_delay)) +
  geom_point(aes(color = origin)) + xlab("Month") + 
  ylab("Average Delay") + geom_smooth(method=loess,color = "Red") +
  ggtitle("NY Airports Average Delays by Month") +
  scale_x_continuous(breaks=1:12, labels=c("Jan","Feb","Mar","Apr","May","Jun",
                                           "Jul","Aug","Sept","Oct","Nov","Dec"))+theme_classic()
  
data1 <- flights %>% group_by(origin, month)  %>%
  summarise(n=n()) %>%
  ungroup() %>%
  arrange(-n)
plot2<-ggplot(data1, aes(x=month, y=n)) +
  geom_point(aes(color=origin)) +
  geom_smooth() +
  ggtitle("Average No. of Flights per Month for NY Airports") +
  scale_x_continuous(breaks=1:12, labels=c("Jan","Feb","Mar","Apr","May","Jun",
                                           "Jul","Aug","Sept","Oct","Nov","Dec"))+ theme_classic()+
 xlab("Month") + ylab("Count")


library(patchwork)
plot1 + plot2
```

As seen above the delays are biggest in June and July, which is the largest period also for flights leaving the three airports. The decrease in delays early in 2013 matches the lower flights numbers, as does the sharp decrease in average delays starting in August to Oct/Nov, then there is a sharp increase in Delays for December which goes against the trend slightly.





Below the percentage of flights that are on-time/delayed is calculated and displayed

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=2.5,  out.width="50%", fig.align='center'}
df<-flightsx4 %>% group_by(duration, time_status) %>%
  summarise(n=n())
#from the dataframe, the percentages can be calculated as:
perc <- c(17.8,82.2,19.7,80.3,22.2,77.8)
#thus plotted
df$percentage_of_flights <- perc
ggplot(data=df,aes(x=time_status,y=percentage_of_flights,fill=percentage_of_flights)) +
  geom_col() +
  facet_wrap(~duration) +
  ggtitle("Percentage of flights delayed/on-time by flight duration")+theme_classic()+
  xlab("Time Status") + ylab("Percentage of Flights")
```


### Planes Analysis

```{r echo=FALSE, message=FALSE, warning=FALSE}
#merge planes and flights by tailnum  and  rename two "year" columns
flights <- flights %>% left_join(planes, by = "tailnum")%>%
  rename(year = year.x) %>%
  rename(manufacture_year = year.y)
```
The reason that looking at amount of seats of planes is as the number of seats can be means more passengers on the plane,sometimes this factor will impact the speed of passengers boarding and disembarking, then lead to the delay of departure.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=3, out.width="50%", fig.align='center'}
# calculate average delay time(departure delay  + arrival delay) group by seats
ave_delay_seats <- flights %>%
                      select(seats,dep_delay,arr_delay)%>%
                      group_by(seats)%>%
                      summarise(ave_delay = mean(dep_delay,na.rm = T) + 
                                  mean(arr_delay,na.rm = T))%>%
                      arrange(-ave_delay)

ggplot(ave_delay_seats,aes(x = seats, y = ave_delay))+
  geom_point()+geom_smooth(method = "lm",color = 'yellow')+ ylab("delay(minutes)")+
  xlab('Number of seats') + theme_classic()+ ggtitle('Delay vs number of seats')
```
  
  
From the graph above, as the amount of seats increases, the delay time shows an upward trend. 

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=3}
# relationship between delay and manufacturer
ave_delay_manufacture <- flights %>%
            select(manufacturer,dep_delay,arr_delay)%>%
            group_by(manufacturer)%>%
            summarise(ave_delay = mean(dep_delay,na.rm = T) + mean(arr_delay,na.rm = T))%>%
            arrange(-ave_delay) %>% na.omit() 
```

Graph of delay time with manufacturers of planes
```{r echo=FALSE,message=FALSE, warning=FALSE,fig.align='center', fig.width=8, fig.height=4,     out.width="70%"}
ggplot(ave_delay_manufacture,aes(x= manufacturer, y = ave_delay,col = manufacturer))+
  geom_point(size = 3)+xlab('Manufacturer')+
  ylab("Delay in Minutes")+
  ggtitle('Delay vs Manufacturer')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position ='none')
```



```{r echo=FALSE, message=FALSE, warning=FALSE,fig.width=6, fig.height=3,out.width="50%",fig.align='center'}
#calculate the age of planes
flights$age <- flights$year - flights$manufacture_year
# calculate average delay time(departure + arrival) group by age of planes
ave_delay_age <- flights %>%
                      select(age,dep_delay,arr_delay,engine)%>%
                      group_by(age,engine)%>%
                      summarise(ave_delay = mean(dep_delay,na.rm = T) + 
                                  mean(arr_delay,na.rm = T))%>%
                      arrange(-ave_delay)%>% na.omit()
ggplot(data = ave_delay_age)+
  geom_point(aes(x= age, y = ave_delay,color = engine),size = 2)+
  geom_smooth(aes(x = age, y = ave_delay),color = 'red')+
  xlim(0,60)+ ylab('Delay in Minutes)')+ 
  ggtitle('Delay vs Age')+
  theme_classic()
```


## Discussion 
```{r}

```


```{r}

```


```{r}

```

